Generalized Universe Hierarchies and First-Class Universe Levels

### 2021-02-21 paper.tex
progress

\section{Universe Features in TTFL}

@@

\textbf{Transfinite hierarchies} are naturally supported. For example, $\Lvl$
can be identified with $\msf{Maybe}\,\Nat_0$, where $\msf{Nothing}$ defines
$\omega$ and $\msf{Just}\,n$ is a finite level. Then, by the definition of
morphisms, we have $<\!\omega : \Pi(n : \Nat_0) \to \msf{Just}\,n <^I \omega$.
We can use this to quantify over finite levels, as in the following type:
\begin{alignat*}{3}
  \Pi(n : \Nat_{\omega})(A : \U\,n\,\omega\,(\mkMor\,(<\!\omega\,n))) \to \Lift\,(\mkMor\,(<\!\omega\,n))\,A \to \Lift\,(\mkMor\,(<\!\omega\,n))\,A
\end{alignat*}

@@

@@

>>>>

\subsection{Universe Features in TTFL}

@@

\textbf{Transfinite hierarchies} are naturally supported. For example, $\Lvl$
can be identified with $\msf{Maybe}\,\Nat_0$, where $\msf{Nothing}$ defines
$\omega$ and $\msf{Just}\,n$ is a finite level. Then, by the definition of
morphisms, we have $<\!\omega : \Pi(n : \Nat_0) \to \msf{Just}\,n <^I \omega$.
We can use this to quantify over finite levels, as in the following type:
\begin{alignat*}{3}
  \Pi(n : \Nat_{\omega})(A : \U\,n\,\omega\,(\mkMor\,(<\!\omega\,n))) \to \Lift\,(\mkMor\,(<\!\omega\,n))\,A \to \Lift\,(\mkMor\,(<\!\omega\,n))\,A
\end{alignat*}
This type is in $\Ty\,\Gamma\,\omega$, but it is not in any universe, since
$\omega$ is the greatest level.

@@

\subsection{Effects of Choice of Level Structure}

TTFL features clearly vary depending on level structures. We make some basic
observations.

\begin{itemize}
\item We did not mandate that the level of $\Lvl^I$ is the least level,
i.e.\ that $l_0 < i$ for every $i \neq l_0$. If this holds, then it is
possible to have level polymorphism at every level: at $l_0$ we can just bind
a $\Lvl^I$, and at every other level, we can lift $\Lvl^I$ to that
level. However, levels are not necessarily totally ordered, and $l_0$ does not
have to be the least. This means that universe polymorphism is prohibited in
levels which are not connected to $l_0$.

\item If levels are given by a limit ordinal, then every TTFL type is contained
in a universe. If levels form a successor ordinal, then this is not the
case. For example, Agda 2.6.1 has $\omega + 1$ levels (externally), where
$\Set{\omega}$ is the topmost universe, but $\Set{\omega}$ has no type.

\item
While it is possible to quantify over all levels (using plain $\Pi$ types), it
is not possible to have level polymorphism over all levels. We may try to type
an identity function for all levels, as $\Pi(i : \Lift\,?\,\Lvl^I)(A :
\U\,(\mkLvl\,i)\,?\,?) \to \Lift\,?\,A \to \Lift\,?\,A$. The issue is in
$\U\,(\mkLvl\,i)\,?\,?$, where we have to find a level which is larger than
\emph{every} level. The solution to this issue is to simply add more levels. For
example, for polymorphism over finite levels, we may pick $\omega + \omega$ as
the first limit ordinal which can internally type finite level polymorphism;
this is what Agda 2.6.2 does.
\end{itemize}

@@

Universes:
Martin-Löf (U in U), Girard, Martin-Löf (countable predicative),
CoC, Coq, Luo ECC. Universe poly in Agda (?), in Coq, cumulativity,
Sterling, Coquand normalization (?)

Transfinite: super, higher-order, Mahlo, IR, indexed IR

Harper univ checking?
